A Bayesian method for latitudinal temperature gradients in the early Eocene

### 2023-02-07 manuscript/manuscript.Rmd
update discussion

# Discussion
The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data. This is an advancement over previously used linear or quadratic approximations [e.g. @Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient, although our generalised logistic function is flexible enough to approximate those shapes, too. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients. Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022], but have the disadvantage of having no model parameters that can be readily compared between data sets. In contrast, the four parameters that describe the shape of our generalised logistic model clearly relate to the magnitude, steepness and the latitude of greatest steepness of the modeled gradient. Given that at least vague prior information on the shape of latitudinal temperature gradients on Earth should exist for any geological time period, we argue that parametric methods should be preferred for reconstructing those gradients.

The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods). A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies. An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).

@@

>>>>

# Discussion
The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data. This is an advancement over previously used linear or quadratic approximations [e.g. @Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient. Our generalised logistic function is flexible enough to approximate the shapes of all of those functions. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients. Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022], but are unsuitable for extrapolation, and have no model parameters that can be readily compared between data sets. When used for interpolation, non-paremetric methods tend to strictly respond to the data, even if the resulting patterns are unlike those predicted by physical processes []. In contrast, the four parameters that describe the shape of our generalised logistic model clearly relate to the magnitude, steepness and the latitude of greatest steepness of the modeled gradient. Given that at least vague prior information on the shape of latitudinal temperature gradients on Earth should exist for any geological time period, we argue that parametric methods should be preferred for reconstructing those gradients.

The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods). In practice, those priors can be informed by physical constraints, e.g. restricting water temperatures to above the freezing point. A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.

@@

An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (maybe cut, otherwise put something to prove that in SM).
